Analysis for Biracial Grade Retention Project

### 2022-09-02 analysis/analysis.Rmd
Add missing variable (state_alaska) to dummied model

```{r counterfactual-ses, warning=FALSE, fig.height=12, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if there was no difference by race and the only compositional differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married). 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
# set up the initial counterfactual dataset
counter_df <- acs %>%
  mutate(year=2014.5, # rough midpoint
         current_grade="6th", # rough midpoint from cumulative probability (55%)
         state="Minnesota", # median state effect in model 4 (Minnesota)
         metro="Metro non-central",
         sex="Male",
         race_real=race,
         race="Black")

coefs <- lapply(unique(acs$race), function(x) {
  means_margin <- marg(model4, "race", data=subset(counter_df, race_real==x))[[1]]
  return(tibble(term=x, estimate=means_margin$Margin, 
                se=means_margin$Standard.Error))
}) %>%
  bind_rows() %>%
  mutate(multiracial=str_detect(term, "/"))

@@

  #now use this as the levels of term as a factor
  coef_table <- coef_table %>%
    mutate(term=factor(term, levels=race_full))

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average adjusted predicted probability of being behind expected grade")

@@

|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
| White/Asian       |       X         |               |         |                |                  |
| Black/Asian       |                 |  X            |         |                |                  |
| Black/Latino      |                 |  X            |         |                |                  |
| White/Latino      |                 |  X            |         |                |                  |
| Indigenous/Latino |                 |  X            |         |                |                  |
| Latino/Asian      |                 |  X            |         |                |                  |
| White/Indigenous  |                 |               |    X    |                |                  |
| Indigenous/Asian  |                 |               |    X    |                |                  |
| Black/Indigenous  |                 |               |    X    |                |                  |
| Black/White       |                 |               |         |         X      |                  |

>>>>

```{r create-dummied-data}
acs_dummied <- acs %>%
  # first we need to unorder current grade for dummy_cols to work
  mutate(current_grade=factor(current_grade, levels=levels(current_grade), 
                              ordered = FALSE)) %>%

  dummy_cols(remove_first_dummy = TRUE) %>%
  select(cluster, strata, perwt, below_exp_grade, year, family_income, 
         parents_married, own_home, degree_mother, degree_father,
         foreign_born, foreign_born_mother, foreign_born_father, 
         spk_eng, spk_eng_father, spk_eng_mother,
         starts_with(c("state_","current_grade_","race","sex_","metro_"))) %>%
  mutate(year=year-2010) %>%
  clean_names()
```

```{r model-dummied}
model_full <- glm(below_exp_grade~
                    year*(current_grade_1st+current_grade_2nd+current_grade_3rd+
                            current_grade_4th+current_grade_5th+
                            current_grade_6th+current_grade_7th+
                            current_grade_8th+current_grade_9th+
                            current_grade_10th+current_grade_11th+
                            current_grade_12th)+
                    race_black+race_indigenous+race_asian+race_latino+
                    race_black_white+race_black_indigenous+race_black_latino+
                    race_black_asian+race_white_indigenous+race_white_latino+
                    race_white_asian+race_indigenous_latino+
                    race_indigenous_asian+race_latino_asian+
                    sex_female+
                    metro_non_metro+metro_central_city+metro_metro_non_central+
                    metro_metro_indeterminable+
                    state_alaska+state_arizona+state_arkansas+state_california+
                    state_colorado+state_connecticut+state_dc+state_delaware+
                    state_florida+state_georgia+state_hawaii+state_idaho+
                    state_illinois+state_indiana+state_iowa+state_kansas+
                    state_kentucky+state_louisiana+state_maine+state_maryland+
                    state_massachusetts+state_michigan+state_minnesota+
                    state_mississippi+state_missouri+state_montana+
                    state_nebraska+state_nevada+state_new_hampshire+
                    state_new_jersey+state_new_mexico+state_new_york+
                    state_north_carolina+state_north_dakota+state_ohio+
                    state_oklahoma+state_oregon+state_pennsylvania+
                    state_rhode_island+state_south_carolina+state_south_dakota+
                    state_tennessee+state_texas+state_utah+state_vermont+
                    state_virginia+state_washington+state_west_virginia+
                    state_wisconsin+state_wyoming+
                    +sqrt(family_income/10000)+parents_married+own_home+
                    degree_mother+degree_father+
                    spk_eng+spk_eng_mother+spk_eng_father+
                    foreign_born_mother+foreign_born_father+foreign_born, 
                  data=acs_dummied, family=binomial)
deviance(model_full)
# FIXME - its a little higher so I am missing something
```

```{r create-cf-df}
# Create the counterfactual dataset where non-resource variables are held
# at their mean. What needs to be set at tis mean?
# - current_grade
# - sex
# - race
# - state
# - metro
# - year
means <- acs_dummied %>%
  select(year, starts_with(c("state_","current_grade_","race_","sex_",
                             "metro_"))) %>%
  lapply(function(x) {rep(mean(x), length(x))}) %>%
  bind_rows()

cf_df <- acs_dummied %>%
  select(cluster, strata, perwt, below_exp_grade, race,family_income, 
         parents_married, own_home, degree_mother, degree_father, foreign_born, 
         foreign_born_mother, foreign_born_father, spk_eng, spk_eng_father, 
         spk_eng_mother) %>%
  bind_cols(means)
```

```{r test-marg, eval=FALSE}
# This code chunk just tests to make sure I get the same value by hand as I get 
# when I do it through the marg command.
temp <- subset(cf_df, race=="Black/White")
lor <- predict(model_full, newdata=temp)
mean(exp(lor)/(1+exp(lor)))
marg(model_full, "year", data=temp)
# It works!
```

```{r create-cf-marg}
coefs <- lapply(unique(acs$race), function(x) {
  temp <- subset(cf_df, race==x)
  means_margin <- marg(model_full, "year", data=temp)[[1]]
  return(tibble(term=x, estimate=means_margin$Margin, 
                se=means_margin$Standard.Error))
}) %>%
  bind_rows() %>%
  mutate(multiracial=str_detect(term, "/"))

@@

  #now use this as the levels of term as a factor
  coef_table <- coef_table %>%
    mutate(term=factor(term, levels=race_full))
```

```{r counterfactual-ses, warning=FALSE, fig.cap="Counterfactual probability of being behind grade for biracial respondents in comparison to their monoracial comparison groups if the only differences between groups were resources (household income, parental education, home ownership, parent and child foreign-born status, parent and child English proficiency, and whether parents are married). 83.4% confidence bands shown around each estimate. Non-overlap in confidence bands roughly indicates statistically significant difference at the 5% level."}
  ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                         ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
    geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                  ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
              color=NA, alpha=0.2)+
    geom_linerange(data=subset(coef_table, multiracial))+
    geom_point()+
    coord_flip()+
    facet_wrap(~mrace, scales="free_y")+
    theme_bw()+
    theme(legend.position = "none", panel.grid = element_blank())+
    scale_color_manual(values=c("grey40","red"))+
    scale_fill_manual(values=c("grey40","red"))+
    scale_linetype_manual(values=c(3,1))+
    scale_y_continuous(labels = scales::percent)+
    labs(x=NULL,
         y="average adjusted predicted probability of being behind expected grade")

@@

|                   | Lower than both | Same as lower | Between | Same as higher | Higher than both |
|-------------------|:---------------:|:-------------:|:-------:|:--------------:|:----------------:|
| White/Asian       |       X         |               |         |                |                  |
| Black/Asian       |                 |  X  (Asian)   |         |                |                  |
| Black/Latino      |                 |  X  (Black)   |         |                |                  |
| White/Latino      |                 |  X  (White)   |         |                |                  |
| Indigenous/Latino |                 |X  (Indigenous)|         |                |                  |
| Latino/Asian      |                 |  X (Asian)    |         |                |                  |
| White/Indigenous  |                 |               |    X    |                |                  |
| Indigenous/Asian  |                 |               |    X    |                |                  |
| Black/Indigenous  |                 |               |    X    |                |                  |
| Black/White       |                 |               |         |   X (Black)    |                  |
